---
title: "Example: PAM50 subtypes in the Mainz dataset"
author: "John Lövrot"
date: "`r format(Sys.Date(), format='%B %d, %Y')`"
output: rmarkdown::tufte_handout
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path='graphics/example-pam50-in-mainz-', 
    echo=FALSE, warning=FALSE, message=FALSE)
```

```{r load_project}
setwd("..")
ProjectTemplate::reload.project()
setwd("reports")
```

```{r settings}
theme_set(theme_bw())

col4 <- rev(RColorBrewer::brewer.pal(4, "RdYlBu"))
names(col4) <- paste0("Q", 1:4)

options(xtable.comment=FALSE)
options(xtable.booktabs=TRUE)
options(xtable.size="footnotesize")
```

Illustrative cluster heatmap
----------------------------

```{r, fig.width=8, fig.height=9, fig.cap="Semi-unsupervised clustering of the Mainz patients based on the PAM50 genes. Average-linkage hierarchical clustering using a one-minus-spearman rank correlation distance metric. LA: Luminal A; LB: Luminal B; H2: HER2-enriched; LB: Basal-like; NBL: Normal breast-like; PAM50PROLIF: PAM50 proliferation index; ER: Estrogen receptor."}
data("pam50", package="genefu")
eset50 <- mainz[fData(mainz)$entrezid %in% pam50$centroids.map$EntrezGene.ID, ]
eset50 <- genefilter::featureFilter(eset50)
featureNames(eset50) <- fData(eset50)$symbol

## Derive relative expression
exprs(eset50) <- sweep(exprs(eset50), 1, apply(exprs(eset50), 1, "median"))

fData(eset50)$PAM50PROLIF <- fData(eset50)$entrezid %in% Nielsen10CCRTabS1$entrezid

fig <- annHeatmap2(
    exprs(eset50),
    dendrogram=list(
        Col=list(clustfun=hclust_avl, distfun=dist_cor, status="yes", lwd=1), 
        Row=list(clustfun=hclust_avl, distfun=dist_cor, status="hide")), 
    annotation=list(
        Col=list(data=pData(eset50)[, c("erstat", "PAM50PROLIF", "subtypecd")], inclRef=TRUE), 
        Row=list(data=fData(eset50)[, "PAM50PROLIF", drop=FALSE], inclRef=FALSE)), 
    labels=list(
        Col=list(labels=NULL),
        Row=list(cex=0.75)),
    breaks=atanbreaks(eset50, trg=4, n=128),
    col=cyanblackyellow,
    legend=TRUE, 
    scale="none")
plot(fig, widths=c(1, 5, 0.5), heights=c(1.5, 5, 2))
```

\clearpage

Assocations with outcome
========================

Illustration of excess distant metastasis plots
-----------------------------------------------

The association between the PAM50 proliferation index and outcome is illustrated using exploratory plots of excess distant metastases. 
A smoother with exploratory confidence band is superimposed in the scatterplot and the contributions from individual patients are shown with circles. 
The shape of the smoother indicates the form of an association between the index and risk of distant metastasis. 
Mathematically, the excess distant metastases are martingale residuals in a null Cox model. 
Corresponding plots for PAM50 intrinsic subtypes are added for comparison.

```{r, fig.width=8, fig.height=4, fig.cap="Excess distant metastases in relation to PAM50 (A) intrinsic subtype and (B) proliferation index."}
ylim0 <- range(mainz$excessdm)

gg1 <- ggplot(data=pData(mainz), aes(x=PAM50PROLIF, y=excessdm)) + 
    geom_point(shape=1, aes(col=cat4(PAM50PROLIF))) + 
    geom_smooth(method="loess", span=0.75, col="black") +
    ylim(ylim0) +
    scale_colour_manual(values=col4, guide=FALSE) + 
    labs(x="Proliferation index", y="Excess distant metastases")

gg2 <- ggplot(data=pData(mainz), aes(x=subtypecd, y=excessdm)) + 
    geom_point(shape=1) + 
    stat_summary(fun.data="mean_cl_normal", size=1.5, aes(col=subtypecd)) +
    scale_colour_manual(values=colCGANsubtypecd, guide=FALSE) + 
    labs(x="Intrinsic subtype", y="Excess distant metastases")

cowplot::plot_grid(gg1, gg2, labels=c("A", "B"), nrow=1)
```

```{r, fig.width=8, fig.height=5.5, fig.cap="Corresponding Kaplan-Meier curves. The PAM50 proliferation index is categorised into quarters."}
layout(matrix(1:2, nrow=1, byrow=TRUE))

survplot(Surv(survdmtm/365.25*12, dmstat) ~ cat4(PAM50PROLIF), 
    data=pData(mainz), 
    col=col4, lwd=2, xmax=7, 
    stitle="Proliferation index", main="A",
    xlab="Time (years)", ylab="Distant metastasis-free survival")

survplot(Surv(survdmtm/365.25*12, dmstat) ~ subtypecd, 
    data=pData(mainz), 
    col=colCGANsubtypecd[levels(mainz$subtypecd)], lwd=2, xmax=7, 
    stitle="Intrinsic subtype", main="B",
    xlab="Time (years)", ylab="Distant metastasis-free survival")

```

\clearpage

Added value of ROR score in ER+ patients
----------------------------------------

Initial exploratory plots:

```{r, fig.width=8, fig.height=4, fig.cap="Excess distant metastases in relation to (A) Nottingham prognostic index and (B) PAM50 risk of reccurence score in ER+ patients. NPI: Nottingham prognostic index (based on tumour size, lymph node status and histological grade); RORS: Risk of reccurence score (subtype alone)."}
ylim0 <- range(mainz$excessdm)

gg1 <- ggplot(
    data=subset(pData(mainz), erstat %in% "ER+"), 
    aes(x=NPI, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", span=1, col="red") +
    facet_grid(. ~ erstat) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

gg2 <- ggplot(
    data=subset(pData(mainz), erstat %in% "ER+"), 
    aes(x=RORS, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", col="red") +
    facet_grid(. ~ erstat) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

cowplot::plot_grid(gg1, gg2, labels=c("A", "B"), nrow=1)
```

Formal statistical inference:

```{r}
model1 <- coxph(Surv(survdmtm, dmstat) ~ NPI, 
    data=subset(pData(mainz), erstat %in% "ER+"))
model2 <- coxph(Surv(survdmtm, dmstat) ~ RORS, 
    data=subset(pData(mainz), erstat %in% "ER+"))
model3 <- coxph(Surv(survdmtm, dmstat) ~ NPI + RORS, 
    data=subset(pData(mainz), erstat %in% "ER+"))

## Concordance index (C-index)
cindex_df <- data.frame(
    marker=factor(c("NPI", "RORS", "NPI+RORS"), 
        levels=c("NPI", "RORS", "NPI+RORS")), 
    cindex=c(
        summary(model1)$concordance["concordance.concordant"], 
        summary(model2)$concordance["concordance.concordant"], 
        summary(model3)$concordance["concordance.concordant"]))

## Likelihood ratio tests
lr_test_df <- rbind(
    cbind(
        data.frame(Comparison="NPI+RORS v RORS"), 
        as.data.frame(anova(model2, model3)[2, 2:4])), 
    cbind(
        data.frame(Comparison="NPI+RORS v NPI"), 
        as.data.frame(anova(model1, model3)[2, 2:4])))
```

```{r, results="asis"}
xtab <- xtable(lr_test_df, digits=3)
caption(xtab) <- "Likelihood ratio tests to assess the added value of ROR-S beyond standard clinical-pathological prognosticators as represented by the Nottingham prognostic index (NPI) in ER+ patients."
print(xtab, include.rownames=FALSE)
```

\vspace{5cm}

```{r, fig.width=4, fig.height=4, fig.margin=TRUE, fig.cap="Concordance indices to assess the added value of ROR-S beyond NPI in ER+ patients."}

colmarker <- colJCO[c("blue", "yellow", "red")]
names(colmarker) <- c("NPI", "RORS", "NPI+RORS")

gg <- ggplot(data=cindex_df, aes(x=marker, y=cindex, fill=marker)) +
    geom_bar(stat="identity", position="dodge") +
    coord_cartesian(ylim=c(0.5, 0.8)) +
    scale_fill_manual(values=colmarker, guide=FALSE) +
    labs(x="", y="C index")

plot(gg)
```
