---
title: "Example: PAM50 subtypes in the Mainz dataset"
author: "John Lövrot"
license: "CC BY 4.0"
date: "`r format(Sys.Date(), format='%B %d, %Y')`"
output: rmarkdown::tufte_handout
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.path='graphics/example-pam50-in-mainz-', 
    echo=FALSE, warning=FALSE, message=FALSE)
```

```{r load_project}
setwd("..")
ProjectTemplate::reload.project()
setwd("reports")
```

```{r settings}
theme_set(theme_bw())

col4 <- rev(RColorBrewer::brewer.pal(4, "RdYlBu"))
names(col4) <- paste0("Q", 1:4)

options(xtable.comment=FALSE)
options(xtable.booktabs=TRUE)
options(xtable.size="footnotesize")
```

```{r, additional_munging}
mainz$trigrp <- with(pData(mainz), factor(
    ifelse(erstat %in% "ER+" & !(subtypecd %in% "H2"), "ER+/not H2", 
        ifelse(erstat %in% "ER-" & !(subtypecd %in% "H2"), "ER-/not H2", 
            ifelse(subtypecd %in% "H2", "H2", NA))), 
    levels=c("ER+/not H2", "H2", "ER-/not H2")))
```

Initial data explorations
=========================

Illustrative cluster heatmap
----------------------------

```{r, fig.width=8, fig.height=9, fig.cap="Semi-unsupervised clustering of the Mainz patients based on the PAM50 genes. Average-linkage hierarchical clustering using a one-minus-spearman rank correlation distance metric. LA: Luminal A; LB: Luminal B; H2: HER2-enriched; LB: Basal-like; NBL: Normal breast-like; PAM50PROLIF: PAM50 proliferation index; ER: Estrogen receptor."}
data("pam50", package="genefu")
eset50 <- mainz[fData(mainz)$entrezid %in% pam50$centroids.map$EntrezGene.ID, ]
eset50 <- genefilter::featureFilter(eset50)
featureNames(eset50) <- fData(eset50)$symbol

## Derive relative expression
exprs(eset50) <- sweep(exprs(eset50), 1, apply(exprs(eset50), 1, "median"))

fData(eset50)$PAM50PROLIF <- fData(eset50)$entrezid %in% Nielsen10CCRTabS1$entrezid

fig <- annHeatmap2(
    exprs(eset50),
    dendrogram=list(
        Col=list(clustfun=hclust_avl, distfun=dist_cor, status="yes", lwd=1), 
        Row=list(clustfun=hclust_avl, distfun=dist_cor, status="hide")), 
    annotation=list(
        Col=list(data=pData(eset50)[, c("erstat", "PAM50PROLIF", "subtypecd")], inclRef=TRUE), 
        Row=list(data=fData(eset50)[, "PAM50PROLIF", drop=FALSE], inclRef=FALSE)), 
    labels=list(
        Col=list(labels=NULL),
        Row=list(cex=0.75)),
    breaks=atanbreaks(eset50, trg=4),
    col=cyanblackyellow,
    legend=TRUE, 
    scale="none")
plot(fig, widths=c(1, 5, 0.5), heights=c(1.5, 5, 2))
```

\clearpage

Assocations with outcome
========================

Illustration of excess distant metastases plots
-----------------------------------------------

The association between the PAM50 proliferation index and outcome is illustrated using exploratory plots of excess distant metastases. 
A smoother with exploratory confidence band is superimposed in the scatterplot and the contributions from individual patients are shown with circles. 
The shape of the smoother indicates the form of an association between the index and risk of distant metastasis. 
Mathematically, the excess distant metastases are martingale residuals in a null Cox model. 
Corresponding plots for PAM50 intrinsic subtypes are added for comparison.

```{r, fig.width=8, fig.height=4, fig.cap="Excess distant metastases in relation to PAM50 (A) intrinsic subtype and (B) proliferation index."}
ylim0 <- range(mainz$excessdm)

gg1 <- ggplot(data=pData(mainz), aes(x=PAM50PROLIF, y=excessdm)) + 
    geom_point(shape=1, aes(col=cat4(PAM50PROLIF))) + 
    geom_smooth(method="loess", span=0.75, col="black") +
    ylim(ylim0) +
    scale_colour_manual(values=col4, guide=FALSE) + 
    labs(x="Proliferation index", y="Excess distant metastases")

gg2 <- ggplot(data=pData(mainz), aes(x=subtypecd, y=excessdm)) + 
    geom_point(shape=1, position=position_jitter(w=0.1, h=0)) + 
    stat_summary(fun.data="mean_cl_normal", size=1.5, aes(col=subtypecd)) +
    scale_colour_manual(values=colCGANsubtypecd, guide=FALSE) + 
    labs(x="Intrinsic subtype", y="Excess distant metastases")

cowplot::plot_grid(gg1, gg2, labels=c("A", "B"), nrow=1)
```

```{r, fig.width=8, fig.height=5.5, fig.cap="Corresponding Kaplan-Meier curves. The PAM50 proliferation index is categorised into quarters."}
layout(matrix(1:2, nrow=1, byrow=TRUE))

survplot(Surv(survdmtm/365.25*12, dmstat) ~ cat4(PAM50PROLIF), 
    data=pData(mainz), 
    col=col4, lwd=2, xmax=7, 
    stitle="Proliferation index", main="A",
    xlab="Time (years)", ylab="Distant metastasis-free survival")

survplot(Surv(survdmtm/365.25*12, dmstat) ~ subtypecd, 
    data=pData(mainz), 
    col=colCGANsubtypecd[levels(mainz$subtypecd)], lwd=2, xmax=7, 
    stitle="Intrinsic subtype", main="B",
    xlab="Time (years)", ylab="Distant metastasis-free survival")

```

\clearpage

Proliferation, intrinsic subtypes and outcome
---------------------------------------------

```{r, fig.width=12, fig.height=6, fig.fullwidth=TRUE, fig.cap="Proliferation, intrinsic subtype and outcome."}
ylim0 <- range(mainz$excessdm)

gg1 <- ggplot(data=pData(mainz), aes(x=PAM50PROLIF, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", span=0.75, col="black") +
    ylim(ylim0) +
    labs(x="Proliferation index", y="Excess distant metastases")

gg2 <- ggplot(data=pData(mainz), aes(x=subtypecd, y=PAM50PROLIF)) + 
    geom_boxplot(aes(fill=subtypecd), outlier.colour="transparent") + 
    geom_point(shape=1, position=position_jitter(w=0.1, h=0)) + 
    scale_x_discrete(limits=rev(levels(mainz$subtypecd))) +
    scale_fill_manual(values=colCGANsubtypecd, guide=FALSE) + 
    labs(x="Intrinsic subtype", y="Proliferation index") + 
    coord_flip()

gg3 <- ggplot(data=pData(mainz), aes(x=PAM50PROLIF, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="lm", aes(col=subtypecd)) +
    ylim(ylim0) +
    facet_wrap(~ subtype) +
    scale_colour_manual(values=colCGANsubtypecd, guide=FALSE) + 
    labs(x="Proliferation index", y="Excess distant metastases")

cowplot::ggdraw() +
    cowplot::draw_plot(gg1, 0, 1/2, 1/4, 1/2) +
    cowplot::draw_plot(gg2, 0, 0, 1/4, 1/2) +
    cowplot::draw_plot(gg3, 1/4, 0, 3/4, 1) +
    cowplot::draw_plot_label(c("A", "B", "C"), x=c(0, 1/4, 0), y=c(1, 1, 1/2))
```

See, for example, the section "Prognostic Signatures Within Intrinsic Subtypes" the review Ades14JCO (PMID [25049332](http://www.ncbi.nlm.nih.gov/pubmed/25049332)).

\clearpage

Added value of ROR score in ER+ patients
----------------------------------------

Initial exploratory plots:

```{r, fig.width=8, fig.height=8, fig.cap="Excess distant metastases in relation to (upper row) Nottingham prognostic index and (lower row) PAM50 risk of reccurence score in (left column) ER+ patients and (right column) ER+/not H2 patients. NPI: Nottingham prognostic index (based on tumour size, lymph node status and histological grade); RORS: Risk of reccurence score (subtype alone)."}
ylim0 <- range(mainz$excessdm)

gg1 <- ggplot(
    data=subset(pData(mainz), erstat %in% "ER+"), 
    aes(x=NPI, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", span=1, col="red") +
    facet_grid(. ~ erstat) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

gg2 <- ggplot(
    data=subset(pData(mainz), trigrp %in% "ER+/not H2"), 
    aes(x=NPI, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", span=1, col="red") +
    facet_grid(. ~ trigrp) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

gg3 <- ggplot(
    data=subset(pData(mainz), erstat %in% "ER+"), 
    aes(x=RORS, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", col="red") +
    facet_grid(. ~ erstat) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

gg4 <- ggplot(
    data=subset(pData(mainz), trigrp %in% "ER+/not H2"), 
    aes(x=RORS, y=excessdm)) + 
    geom_point(shape=1) + 
    geom_smooth(method="loess", col="red") +
    facet_grid(. ~ trigrp) +
    ylim(ylim0) +
    labs(y="Excess distant metastases")

cowplot::plot_grid(gg1, gg2, gg3, gg4, 
    labels=c("A", "B", "C", "D"), nrow=2)
```

\clearpage

Formal statistical inference, compare with, for example, Dowsett13JCO (PMID [23816962](http://www.ncbi.nlm.nih.gov/pubmed/23816962)):

```{r}
model1a <- coxph(Surv(survdmtm, dmstat) ~ NPI, 
    data=subset(pData(mainz), erstat %in% "ER+"))
model2a <- coxph(Surv(survdmtm, dmstat) ~ RORS, 
    data=subset(pData(mainz), erstat %in% "ER+"))
model3a <- coxph(Surv(survdmtm, dmstat) ~ NPI + RORS, 
    data=subset(pData(mainz), erstat %in% "ER+"))

model1b <- coxph(Surv(survdmtm, dmstat) ~ NPI, 
    data=subset(pData(mainz), trigrp %in% "ER+/not H2"))
model2b <- coxph(Surv(survdmtm, dmstat) ~ RORS, 
    data=subset(pData(mainz), trigrp %in% "ER+/not H2"))
model3b <- coxph(Surv(survdmtm, dmstat) ~ NPI + RORS, 
    data=subset(pData(mainz), trigrp %in% "ER+/not H2"))

## Concordance index (C-index)
cindex_df <- data.frame(
    subpop=factor(rep(c("ER+", "ER+/not H2"), each=3), 
        levels=c("ER+", "ER+/not H2")), 
    marker=factor(rep(c("NPI", "RORS", "NPI+RORS"), times=2), 
        levels=c("NPI", "RORS", "NPI+RORS")), 
    cindex=c(
        summary(model1a)$concordance["concordance.concordant"], 
        summary(model2a)$concordance["concordance.concordant"], 
        summary(model3a)$concordance["concordance.concordant"],
        summary(model1b)$concordance["concordance.concordant"], 
        summary(model2b)$concordance["concordance.concordant"], 
        summary(model3b)$concordance["concordance.concordant"]))

## Likelihood ratio tests
lr_test_df <- rbind(
    cbind(
        data.frame(
            Population=c("ER+"), 
            Comparison="NPI+RORS v RORS"), 
        as.data.frame(anova(model2a, model3a)[2, 2:4])), 
    cbind(
        data.frame(
            Population=c("ER+"), 
            Comparison="NPI+RORS v NPI"), 
        as.data.frame(anova(model1a, model3a)[2, 2:4])),
    cbind(
        data.frame(
            Population=c("ER+/not H2"), 
            Comparison="NPI+RORS v RORS"), 
        as.data.frame(anova(model2b, model3b)[2, 2:4])), 
    cbind(
        data.frame(
            Population=c("ER+/not H2"), 
            Comparison="NPI+RORS v NPI"), 
        as.data.frame(anova(model1b, model3b)[2, 2:4])))
```

```{r, fig.width=8, fig.height=4, fig.margin=FALSE, fig.cap="Concordance indices to assess the added value of ROR-S beyond standard clinical-pathological prognosticators as represented by the Nottingham prognostic index (NPI) in (A) ER+ patients and (B) ER+/not H2 patients."}

colmarker <- colJCO[c("blue", "yellow", "red")]
names(colmarker) <- c("NPI", "RORS", "NPI+RORS")

gg1 <- ggplot(data=subset(cindex_df, subpop == "ER+"), 
    aes(x=marker, y=cindex, fill=marker)) +
    geom_bar(stat="identity", position="dodge") +
    coord_cartesian(ylim=c(0.5, 0.8)) +
    facet_grid(. ~ subpop) +
    scale_fill_manual(values=colmarker, guide=FALSE) +
    labs(x="", y="C index")

gg2 <- ggplot(data=subset(cindex_df, subpop == "ER+/not H2"), 
    aes(x=marker, y=cindex, fill=marker)) +
    geom_bar(stat="identity", position="dodge") +
    coord_cartesian(ylim=c(0.5, 0.8)) +
    facet_grid(. ~ subpop) +
    scale_fill_manual(values=colmarker, guide=FALSE) +
    labs(x="", y="C index")

cowplot::plot_grid(gg1, gg2, labels=c("A", "B"), nrow=1)
```

```{r, results="asis"}
xtab <- xtable(lr_test_df, digits=3)
caption(xtab) <- "Likelihood ratio tests to assess the added value of ROR-S beyond NPI in the sub-populations ER+ patients and ER+/not H2 patients."
print(xtab, include.rownames=FALSE)
```
